
---
title: "Exploring the Central Limit Theorem with the Exponential Distribution"
author: "tjmb03"
date: "2025-07-30"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
set.seed(1234)
```

## Overview

This report demonstrates the Central Limit Theorem (CLT) using the exponential distribution. Specifically, I simulate the distribution of the mean of 40 exponential random variables with a rate parameter lambda = 0.2, repeated 1000 times. We compare the empirical sample statistics with theoretical values and visually assess the normality of the resulting distribution of sample means.

## Simulations

```{r}
lambda <- 0.2
n <- 40
simulations <- 1000

# Simulate 1000 means of 40 exponentials each
exp_means <- replicate(simulations, mean(rexp(n, rate = lambda)))
```

## Sample Mean versus Theoretical Mean

```{r}
sample_mean <- mean(exp_means)
theoretical_mean <- 1 / lambda

cat("Sample Mean:", sample_mean, "\n")
cat("Theoretical Mean:", theoretical_mean, "\n")
```

### Histogram: Distribution of Sample Means with Mean Lines

```{r}
hist(exp_means, probability = TRUE, breaks = 30, col = "lightblue",
     main = "Histogram of Sample Means (40 Exponentials, 1000 Simulations)",
     xlab = "Sample Mean")
abline(v = sample_mean, col = "blue", lwd = 2)
abline(v = theoretical_mean, col = "red", lwd = 2, lty = 2)
legend("topright", legend = c("Sample Mean", "Theoretical Mean"),
       col = c("blue", "red"), lty = c(1, 2), lwd = 2)
```

## Sample Variance versus Theoretical Variance

```{r}
sample_var <- var(exp_means)
theoretical_var <- (1 / lambda)^2 / n

cat("Sample Variance:", sample_var, "\n")
cat("Theoretical Variance:", theoretical_var, "\n")
```

## Distribution: Normality Assessment

```{r}
# Histogram with normal curve
hist(exp_means, probability = TRUE, breaks = 30, col = "lightblue",
     main = "Distribution of Averages of 40 Exponentials",
     xlab = "Average Value")
curve(dnorm(x, mean = mean(exp_means), sd = sd(exp_means)), 
      col = "red", lwd = 2, add = TRUE)

# QQ Plot
qqnorm(exp_means, main = "QQ Plot of Averages of 40 Exponentials")
qqline(exp_means, col = "red")
```

## Motivating Example: Uniform Distribution Comparison

```{r}
# Raw uniforms
uniform_vals <- runif(1000)

# Averages of 40 uniforms
uniform_means <- replicate(1000, mean(runif(40)))

# Histograms
hist(uniform_vals, probability = TRUE, breaks = 30, col = "lightgreen",
     main = "Histogram of 1000 Random Uniform(0,1) Values", xlab = "Value")

hist(uniform_means, probability = TRUE, breaks = 30, col = "lightcoral",
     main = "Histogram of Averages of 40 Uniform(0,1) Values", xlab = "Average Value")
```

## Final Summary

- The Central Limit Theorem is verified using the exponential distribution.
- Even though the exponential distribution is skewed, the distribution of sample means approximates a normal distribution.
- Sample mean and variance match theoretical expectations.
- The same trend is observed with uniform distributions, reinforcing the CLT.
