
---
title: "ToothGrowth Data Analysis"
author: "tjmb03"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

In this report, we analyze the `ToothGrowth` dataset from the `datasets` package in R. The data records the effect of Vitamin C on tooth growth in guinea pigs, using different supplement types (`OJ` and `VC`) and dose levels (0.5, 1.0, 2.0 mg/day).

## Load and Inspect Data

```{r}
# Load data
data("ToothGrowth")
head(ToothGrowth)
str(ToothGrowth)
summary(ToothGrowth)

# Convert dose to a factor
ToothGrowth$dose <- as.factor(ToothGrowth$dose)
```

## Exploratory Data Analysis

### Tooth Length by Supplement Type

```{r fig.width=4, fig.height=3}
library(ggplot2)

ggplot(ToothGrowth, aes(x = supp, y = len, fill = supp)) +
  geom_boxplot() +
  labs(title = "Tooth Length by Supplement Type", x = "Supplement", y = "Tooth Length") +
  theme_minimal()
```

### Tooth Length by Dose

```{r fig.width=4, fig.height=3}
ggplot(ToothGrowth, aes(x = dose, y = len, fill = dose)) +
  geom_boxplot() +
  labs(title = "Tooth Length by Dose", x = "Dose (mg/day)", y = "Tooth Length") +
  theme_minimal()
```

### Tooth Length by Supplement and Dose

```{r fig.width=4, fig.height=3}
ggplot(ToothGrowth, aes(x = dose, y = len, fill = supp)) +
  geom_boxplot(position = position_dodge()) +
  labs(title = "Tooth Length by Supplement and Dose", x = "Dose (mg/day)", y = "Tooth Length") +
  theme_minimal()
```

## Hypothesis Testing

### 1. Supplement Comparison (OJ vs VC)

```{r}
# Test Equal Variance Between supp (OJ vs VC)
var.test(len ~ supp, data = ToothGrowth)
#run t-test
t_supp <- t.test(len ~ supp, data = ToothGrowth, var.equal = TRUE)
t_supp
```

**95% Confidence Interval**:
```{r}
t_supp$conf.int
```

### 2. Dose Comparisons (0.5, 1.0, 2.0)

```{r}
##Test Equal Variance Between Dose Groups (pairwise)
# Dose 0.5 vs 1.0
var.test(len ~ dose, data = subset(ToothGrowth, dose %in% c(0.5, 1.0)))
# Dose 1.0 vs 2.0
var.test(len ~ dose, data = subset(ToothGrowth, dose %in% c(1.0, 2.0)))
# Dose 0.5 vs 2.0
var.test(len ~ dose, data = subset(ToothGrowth, dose %in% c(0.5, 2.0)))

# Pairwise t-tests
pairwise.t.test(ToothGrowth$len, ToothGrowth$dose, p.adj = "none")
```

#### Manual Confidence Intervals

```{r}
t_05_10 <- t.test(len ~ dose, data = subset(ToothGrowth, dose %in% c(0.5, 1.0)), var.equal = TRUE)
t_10_20 <- t.test(len ~ dose, data = subset(ToothGrowth, dose %in% c(1.0, 2.0)), var.equal = TRUE)
t_05_20 <- t.test(len ~ dose, data = subset(ToothGrowth, dose %in% c(0.5, 2.0)), var.equal = TRUE)

t_05_10$conf.int
t_10_20$conf.int
t_05_20$conf.int
```

## Conclusions

- There is a difference with no significance (p=0.06, 95% Confidence Interval) in tooth length between supplement types: **OJ** tends to result in higher tooth length than **VC**.
- There is also a statistically significant difference in tooth length among all doses. Higher doses lead to longer tooth growth.
- All three pairwise dose comparisons showed significant differences, supported by t-tests and 95% confidence intervals.

## Assumptions

- Independent and random sampling
- Approximate normality (reasonable due to sample size)
- Equal variances are handled using standard t-test
